---
title: "Panel Dataa Models"
author: "Peter Kariuki"
date: "7/14/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,comment = F,warning = F,message = F,results = 'asis')
```

## Recap
Difference -in-differences models involves before and after comparisos in natural experiments to determine the effect of a treatment.
It is used in *quasi-experiments*, i.e when treatment assignment is not random



Difference-in-Difference Model:
Y=B0+B1*Time+B2*Treatment+B3*Time*Treatment+u

Estimating regression at baseline, Time=0
Y=B0+B2*Treatment+u
This shows the difference in outcomes between the treated and control units at baseline.

Estimating regression in at the endline, Time=1
Y=B0+B2*Time+B3*Treatment+u

The coefficient (B3) shows the difference in outcomes between the treated and the control units at the endline

The DID effect is the difference between the two differences above, i.e 
(B0+B2+B3)-(B0+B2)
This is the difference in outcomes between treated and control units at baseline and the treated and control units at the endline.

```{r}
# Install packages
PackageNames <- c("tidyverse", "stargazer", "magrittr", "haven")
for(i in PackageNames){
  if(!require(i, character.only = T)){
    install.packages(i, dependencies = T)
    require(i, character.only = T)
  }
}
```

Parallel Trends Assumption

```{r,echo=FALSE}
# initialize plot and add control group
plot(c(0, 1), c(6, 8), 
     type = "p",
     ylim = c(5, 12),
     xlim = c(-0.5, 1.5),
     main = "The Differences-in-Differences Estimator",
     xlab = "Period",
     ylab = "Y",
     col = "steelblue",
     pch = 20,
     xaxt = "n",
     yaxt = "n")

axis(1, at = c(0, 1), labels = c("Baseline", "Endline"))
axis(2, at = c(0, 13))

# add treatment group
points(c(0, 1, 1), c(7, 9, 11), 
       col = "darkred",
       pch = 20)

# add line segments
lines(c(0, 1), c(7, 11), col = "darkred")
lines(c(0, 1), c(6, 8), col = "steelblue")
lines(c(0, 1), c(7, 9), col = "darkred", lty = 2)
lines(c(1, 1), c(9, 11), col = "black", lty = 2, lwd = 2)

# add annotations
text(1, 10, expression(hat(beta)[3]^{DID}), cex = 0.8, pos = 4)
text(0, 5.5, "E[Y|Treatment=Control]", cex = 0.8 , pos = 4)
text(0, 6.8, "E[Y|Treatment=Treatment]", cex = 0.8 , pos = 4)
text(1, 7.9, "E[Y|Treatment=Control]", cex = 0.8 , pos = 4)
text(1, 11.1, "E[Y|Treatment=Treatment]", cex = 0.8 , pos = 4)
```




```{r}
# House prices example
# DID effect of building an incinerator on house prices
directory <-here::here("6. Panel Data Models")
KIELMC <- read.csv("KIELMC.csv")

# Explore variables of interest
KIELMC %>% 
  select(rprice, nearinc,y81,y81nrinc) %>% 
  stargazer(type = "html")

```

```{r}
# Summarize house prices near and far from incinerator, and before and after
KIELMC %>% 
  group_by(nearinc, y81) %>%
  summarize_at(.vars = vars(rprice), 
               .funs = list(mean = mean, obs = length)) %>% 
  kableExtra::kable() %>% kableExtra::kable_styling("hover",full_width = F)
```

```{r}
# Regressionat endline (after building the incinerator)
model1 <- lm(formula = rprice ~ nearinc, 
             data = KIELMC, 
             subset = (year == 1981))
#summary(model1)
(b1 <- coef(model1)["nearinc"])
```

```{r}
# Regression at baseline (before building the incinerator)
model2 <- update(model1, subset = (year == 1978))
#summary(model2)
(b2 <- coef(model2)["nearinc"])
```

```{r}
# Difference-in-differences effect
b1 - b2
```

```{r}
# Difference-in-differences regression 
# includes treated, after, and after*treated
model3 <- lm(rprice ~ nearinc + y81 + y81*nearinc, KIELMC)
#summary(model3)
coef(model3)["nearinc:y81"]
# DID effect is the coefficient on after*treated
# DID effect is same as the difference-in-differences calculated above
```

```{r}
stargazer(model1,model2,model3,type="html",align = T, covariate.labels=c("Price Near Incinerator","Year","Interaction","Constant"))

library(equatiomatic)
extract_eq(model3,intercept = "beta",wrap = TRUE, terms_per_line = 2,ital_vars = TRUE)

extract_eq(model3,intercept = "beta",wrap = TRUE, terms_per_line = 2,ital_vars = TRUE,use_coefs = TRUE)
```

